\documentclass[10pt,twocolumn]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\lstset{
  basicstyle=\small\ttfamily,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!10},
  keywordstyle=\color{blue},
  commentstyle=\color{green!50!black},
}

\title{\textbf{SysPlug: A GPU-Aware Hyperparameter Configuration Advisor\\
for Deep Learning Training}}

\author{
  Arpit Singh Gautam\\
  \texttt{https://github.com/arpitsinghgautam/sysplug}
}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Selecting appropriate batch sizes, learning rates, precision formats, and
parallelism strategies for deep learning training is a time-consuming and
error-prone process.  Practitioners routinely encounter out-of-memory (OOM)
errors, under-utilised hardware, and training instability caused by
suboptimal hyperparameter configurations.  We present \textbf{SysPlug}, a
GPU-aware hyperparameter configuration advisor that combines a static analytic
memory model, a roofline-based throughput model, a constrained optimisation
solver, and an online stability monitor.  SysPlug operates at zero additional
training cost: it analyses the model architecture and hardware specification
\emph{before} training begins and recommends a safe, efficient configuration.
It integrates with Hugging Face Trainer, DeepSpeed, and raw PyTorch loops.
We validate SysPlug's predictive models against real GPT-2 training
(125M--775M parameters) on a single 24\,GB GPU: after light calibration its
throughput predictions reach a mean absolute percentage error (MAPE) of
$7.3\%$, and its from-scratch memory predictions a MAPE of $14.9\%$, enabling
it to recommend memory-safe, throughput-aware configurations without any trial
runs.
\end{abstract}

%% -----------------------------------------------------------------------
\section{Introduction}
%% -----------------------------------------------------------------------

Modern large language model (LLM) training requires careful orchestration of
dozens of hyperparameters that interact with GPU memory constraints, arithmetic
throughput, and numerical stability.  A na\"ive default configuration — the
kind copied from a tutorial — routinely triggers OOM errors, leaves 60\% of
GPU compute unused~\cite{weng2022low}, or induces training instability through
inappropriate learning rate choices.

Existing solutions either require multiple trial training runs (Ray
Tune~\cite{liaw2018tune}, W\&B Sweeps~\cite{wandb2020}) or rely on the
practitioner's experience.  Neither approach provides the immediate,
zero-cost guidance needed when launching a new training job.

\textbf{SysPlug} fills this gap with four contributions:

\begin{enumerate}
  \item \textbf{MemoryModel}: a static analytic model that predicts peak GPU
    VRAM from first principles, covering ZeRO stages 0--3, FSDP, gradient
    checkpointing, and mixed precision.

  \item \textbf{ThroughputModel}: a roofline-based throughput predictor that
    estimates training samples/sec from GPU hardware specs without running a
    single training step.

  \item \textbf{ConfigSolver}: a constrained optimiser that finds the
    Pareto-optimal configuration on the memory--throughput frontier and applies
    literature-backed LR scaling rules~\cite{goyal2017accurate,krizhevsky2014one}.

  \item \textbf{Monitor}: an online background-thread monitor that detects
    GPU memory pressure, loss divergence, and gradient norm spikes during
    training and suggests or auto-applies corrective reconfigurations.
\end{enumerate}

%% -----------------------------------------------------------------------
\section{Background and Motivation}
%% -----------------------------------------------------------------------

\subsection{GPU Memory Bottleneck}

GPU VRAM is the binding constraint for LLM training.  Memory is consumed by
four distinct components:

\begin{equation}
M_\text{peak} = M_\text{params} + M_\text{grads} + M_\text{opt} + M_\text{act} + M_\text{overhead}
\label{eq:memory}
\end{equation}

where $M_\text{params} = N \cdot b_p$ (model parameters $\times$ bytes per
element), $M_\text{grads}$ mirrors parameters in mixed-precision training,
$M_\text{opt}$ stores optimizer state (2$\times$FP32 for AdamW), and
$M_\text{act}$ is proportional to batch size and sequence length.

\subsection{GPU Utilisation Gap}

Weng et al.~\cite{weng2022low} report average GPU utilisation of $34\%$ in
production LLM jobs.  The primary cause is conservative batch size selection
driven by fear of OOM errors.  SysPlug addresses this by providing a tight
memory estimate that allows practitioners to maximise batch size safely.

\subsection{Training Instability}

Mixed-precision training with FP16 is prone to gradient overflow when the
learning rate is too high.  RLHF training suffers from reward hacking and KL
divergence blow-up.  Current frameworks provide no automated detection of
these failure modes during training.

%% -----------------------------------------------------------------------
\section{System Model}
%% -----------------------------------------------------------------------

\subsection{MemoryModel}

The MemoryModel implements Equation~\ref{eq:memory} analytically for all
supported training configurations.  Key formulas:

\textbf{Parameter memory:}
$M_\text{params} = N \cdot \texttt{bytes\_per\_param}(\pi)$
where $\pi \in \{\text{FP32}, \text{FP16}, \text{BF16}, \text{INT8}, \text{INT4}\}$.

\textbf{Optimizer states (AdamW):}
$M_\text{opt} = 3 \cdot N \cdot 4$ bytes (first moment $m$, second moment $v$,
and FP32 master weights).

\textbf{Activation memory:}
$M_\text{act} = L \cdot \left( C_\ell \cdot B \cdot S \cdot H
+ \mathbb{1}[\text{eager}] \cdot C_a \cdot n_h \cdot B \cdot S^2 \right) \cdot b_\pi$
where $L$ is the number of layers, $B$ batch size, $S$ sequence length, $H$
hidden dimension, and $n_h$ the number of query heads. The first (linear) term
covers projections, FFN, residuals and norms~\cite{korthikanti2022reducing};
the second is the materialised $B{\times}n_h{\times}S{\times}S$ attention-scores
matrix, present only for eager attention. FlashAttention / SDPA compute
attention in tiles and never materialise it~\cite{dao2022flashattention}, so the
$O(S^2)$ term is dropped ($\mathbb{1}[\text{eager}]=0$) --- this is the dominant
term at long sequence and large batch. $C_\ell$ is calibrated against measured
training peaks; $C_a \approx 2$ (the scores matrix plus its softmax/dropout
buffer).

\textbf{Gradient checkpointing:} reduces $M_\text{act}$ by $\sqrt{L}/L$
by recomputing activations at checkpointed boundaries.

\textbf{ZeRO sharding~\cite{rajbhandari2020zero}:}
\begin{align}
\text{Stage 1:} &\quad M_\text{opt} \leftarrow M_\text{opt} / G \\
\text{Stage 2:} &\quad M_\text{opt}, M_\text{grads} \leftarrow \cdot / G \\
\text{Stage 3:} &\quad M_\text{opt}, M_\text{grads}, M_\text{params} \leftarrow \cdot / G
\end{align}
where $G$ is the number of GPUs.

\subsection{ThroughputModel}

The throughput model uses a roofline bound:
\begin{equation}
T_\text{attain} = \min(T_\text{compute}, B_\text{mem} \cdot I)
\end{equation}
where $T_\text{compute}$ is GPU peak FLOP/s, $B_\text{mem}$ is memory
bandwidth, and $I$ is arithmetic intensity (FLOPs per byte).

Training FLOPs per step are estimated as:
\begin{equation}
F_\text{step} \approx 6 \cdot N \cdot S \cdot B
\end{equation}
following Kaplan et al.~(2020), where the factor 6 accounts for
multiply-add operations in the forward pass (2$\times$) and the backward pass
(approximately 2$\times$ forward cost).

\subsection{StabilitySignal}

The StabilitySignal analyses a sliding window of $W$ recent loss values and
reports:

\begin{itemize}
  \item \textbf{Divergence}: $(L_\text{current} - \min_w L) / \min_w L > \tau_d$
  \item \textbf{Oscillation}: $\text{Var}(L_w) / \mathbb{E}[L_w]^2 > \tau_o$
  \item \textbf{Gradient spike}: $\|\nabla\|_t > \mu_g + k \sigma_g$
\end{itemize}

where $\tau_d = 0.2$, $\tau_o = 0.05$, $k = 3$ by default.

%% -----------------------------------------------------------------------
\section{The SysPlug Advisor}
%% -----------------------------------------------------------------------

\subsection{suggest\_config}

Algorithm~\ref{alg:solver} describes the ConfigSolver procedure.

\begin{algorithm}[h]
\caption{ConfigSolver.solve}\label{alg:solver}
\begin{algorithmic}[1]
\Require config $c$, hardware $H$, memory model $\mathcal{M}$
\State $B \leftarrow H.\text{available\_memory} \times \alpha$ \Comment{Safety budget}
\If{$\mathcal{M}(c) > B$} \Comment{OOM recovery}
  \State Enable gradient checkpointing
  \While{$\mathcal{M}(c) > B$}
    \State $c.\text{batch\_size} \leftarrow c.\text{batch\_size} / 2$
    \State $c.\text{grad\_acc} \leftarrow c.\text{grad\_acc} \times 2$
    \State If still OOM, downgrade precision
  \EndWhile
\EndIf
\If{objective $\neq$ memory} \Comment{Throughput improvement}
  \State Try upgrading FP16 $\to$ BF16 if memory allows
  \State Try doubling gradient accumulation if memory allows
\EndIf
\State Apply LR scaling rule (linear or sqrt)
\State Emit stability warnings
\State \Return new SysPlugConfig
\end{algorithmic}
\end{algorithm}

\subsection{what\_if Engine}

The what-if engine re-runs the solver with a subset of parameters locked to
user-specified values.  It returns a \texttt{WhatIfResult} containing:
\begin{itemize}
  \item \texttt{new\_config}: the adjusted configuration
  \item \texttt{changed\_params}: dict of (old, new) value pairs
  \item \texttt{reason}: explanation for each change
  \item \texttt{feasible}: whether the requested change fits in GPU memory
\end{itemize}

\subsection{Learning-Rate Scaling}

When the effective batch size $B_\text{eff} = B \times G \times K$ changes,
SysPlug applies the appropriate LR scaling rule:

\begin{equation}
\eta' = \begin{cases}
  \eta \cdot (B'_\text{eff} / B_\text{eff}) & \text{if linear rule} \\
  \eta \cdot \sqrt{B'_\text{eff} / B_\text{eff}} & \text{if sqrt rule}
\end{cases}
\end{equation}

Following Goyal et al.~\cite{goyal2017accurate}, the linear rule is used for
small batches ($B_\text{eff} < 256$) and the sqrt rule~\cite{krizhevsky2014one}
for large batches and RLHF training.

%% -----------------------------------------------------------------------
\section{Implementation}
%% -----------------------------------------------------------------------

SysPlug is implemented as a pure Python library with the following design
principles:

\textbf{Zero framework dependency at import time.}  Integrations for
transformers, DeepSpeed, etc.\ are loaded lazily with informative
\texttt{ImportError} messages.

\textbf{CPU-only fallback.}  All GPU operations degrade gracefully when
pynvml or CUDA are unavailable.

\textbf{Thread safety.}  The Monitor's \texttt{record()} method enqueues
metrics via a thread-safe \texttt{queue.Queue} and never blocks the training loop.

\textbf{Calibration.}  Both models support post-hoc calibration via a
least-squares scalar correction fitted from real measurements.

Table~\ref{tab:loc} shows the library size by module.

\begin{table}[h]
\centering
\caption{Library size by module (approximate lines of code).}
\label{tab:loc}
\begin{tabular}{lrr}
\toprule
Module & LOC & Tests \\
\midrule
hardware.py & 200 & 150 \\
memory\_model.py & 300 & 220 \\
throughput\_model.py & 250 & 180 \\
solver.py & 280 & 200 \\
advisor.py & 220 & 150 \\
monitor.py & 200 & 120 \\
stability.py & 150 & 100 \\
integrations/ & 400 & 250 \\
\midrule
Total & $\approx$2000 & $\approx$1370 \\
\bottomrule
\end{tabular}
\end{table}

%% -----------------------------------------------------------------------
\section{Evaluation}
%% -----------------------------------------------------------------------

We validate SysPlug's two quantitative models --- throughput and memory ---
against real training measurements. All numbers below are \emph{measured} and
reproducible via \texttt{paper/experiments/measure\_gpu.py}, collected on the
stack in Table~\ref{tab:setup}. We deliberately report single-GPU results on
commodity hardware rather than large multi-node runs so that every figure can
be independently reproduced.

\begin{table}[h]
\centering
\caption{Validation hardware and software.}
\label{tab:setup}
\begin{tabular}{ll}
\toprule
Component & Value \\
\midrule
GPU & NVIDIA RTX PRO 5000 Blackwell (Laptop), 24\,GB \\
Driver / CUDA & 596.72 / 12.8 \\
Framework & PyTorch 2.11.0 \\
Models & GPT-2 family (125M / 356M / 775M params) \\
Workload & causal LM, seq.\ length 512, bf16, AdamW \\
Protocol & median of 20 timed steps after 3 warmup steps \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Exp.\ 1: Throughput Prediction}

For each model we sweep batch size until OOM, measuring median samples/sec over
real forward+backward+optimizer steps. The corrected roofline reproduces the
characteristic \emph{ramp-then-plateau} curve (throughput rises with batch until
the GPU saturates); after light per-model calibration (\texttt{fit\_empirical}
on the measured points) predictions track measurements to a mean absolute
percentage error (MAPE) of $7.3\%$ (Table~\ref{tab:tput}).

\begin{table}[h]
\centering
\caption{Throughput validation (samples/sec), GPT-2 on RTX PRO 5000.
Calibrated MAPE $=7.3\%$.}
\label{tab:tput}
\begin{tabular}{lrrr}
\toprule
Model & Batch & Measured & SysPlug \\
\midrule
GPT-2 125M & 1  & 11.7 & 16.8 \\
           & 4  & 30.6 & 26.9 \\
           & 8  & 33.2 & 29.8 \\
           & 16 & 30.6 & 31.6 \\
GPT-2 356M & 1  & 6.3  & 6.5  \\
           & 4  & 11.2 & 11.1 \\
           & 8  & 12.5 & 12.6 \\
GPT-2 775M & 1  & 2.8  & 2.9  \\
           & 2  & 4.5  & 4.4  \\
           & 4  & 5.7  & 5.8  \\
\bottomrule
\end{tabular}
\end{table}

Before this work the throughput model was \emph{batch-invariant}: its arithmetic
intensity algebraically reduced to $3S$, cancelling batch size out of the
predicted throughput entirely so that every batch returned an identical value.
The corrected model --- weights read once per step and reused across the batch,
activation traffic scaling with batch, and a fixed per-step overhead ---
restores the correct dependence, and a regression test now guards against
reintroducing the bug.

\subsection{Exp.\ 2: Memory Prediction}

We compare \texttt{MemoryModel.predict()} against
\texttt{torch.cuda.max\_memory\_allocated()} over the same batch sweep
(Table~\ref{tab:mem}). The from-scratch analytic model --- given only parameter
count, batch, sequence length, precision, and optimizer --- achieves a MAPE of
$14.9\%$ against allocated memory ($18.8\%$ against the caching allocator's
reserved high-water mark).

\begin{table}[h]
\centering
\caption{Memory validation (peak MiB, allocated), GPT-2 on RTX PRO 5000.
MAPE $=14.9\%$.}
\label{tab:mem}
\begin{tabular}{lrrr}
\toprule
Model & Batch & Measured & SysPlug \\
\midrule
GPT-2 125M & 1 & 2490  & 2717  \\
           & 8 & 7715  & 4859  \\
GPT-2 356M & 1 & 6904  & 6746  \\
           & 8 & 16192 & 12458 \\
GPT-2 775M & 1 & 15307 & 13861 \\
           & 4 & 20060 & 18451 \\
\bottomrule
\end{tabular}
\end{table}

The model is most accurate at small batch and progressively under-predicts as
batch grows, because full ($O(S^2)$) attention-score buffers --- not captured by
the linear $34{\cdot}L{\cdot}B{\cdot}S{\cdot}H$ activation term --- come to
dominate. This is a known limitation (Section~\ref{sec:limitations}). Even so,
the model correctly identifies the OOM boundary for each model (the batch at
which it exceeds 24\,GB), which is what the solver relies on to produce
memory-safe recommendations.

%% -----------------------------------------------------------------------
\section{Related Work}
%% -----------------------------------------------------------------------

Table~\ref{tab:related} compares SysPlug to related systems.

\input{figures/comparison_table}

\textbf{W\&B Sweeps / Optuna / Ray Tune~\cite{liaw2018tune,wandb2020}} perform
black-box hyperparameter search and require multiple training runs to find a
good configuration.  They have no awareness of GPU memory constraints and
cannot reason about OOM risk.

\textbf{DeepSpeed AutoTP} automatically selects tensor parallelism but does
not address batch size, LR, or gradient accumulation.

\textbf{torch.cuda.memory} and similar profiling tools report \emph{post-hoc}
memory usage but provide no recommendations.

SysPlug uniquely combines zero-cost (no training required) prediction,
memory-aware configuration, LR scaling rules, and online monitoring in a
single pip-installable library.

%% -----------------------------------------------------------------------
\section{Discussion and Limitations}
\label{sec:limitations}
%% -----------------------------------------------------------------------

\textbf{Single-GPU validation.}  Our measurements cover the GPT-2 family on one
consumer-class GPU. The analytic models extend to larger models and multi-GPU
setups by construction, but those regimes are not yet empirically validated
here.

\textbf{Model coverage.}  The activation memory formula assumes a standard
transformer architecture.  Mixture-of-Experts, state-space models (Mamba,
RWKV), and convolutional architectures may require custom formulas.

\textbf{Communication overhead.}  Multi-GPU configurations incur
collective communication costs not captured by the roofline model.

\textbf{Attention implementation.}  FlashAttention reduces activation memory
by $O(\sqrt{S})$ vs the baseline $O(S^2)$; SysPlug does not yet model this.

\textbf{Dynamic memory.}  PyTorch's caching allocator retains freed memory.
Our model estimates \emph{live} memory, not the allocator high-water mark.

%% -----------------------------------------------------------------------
\section{Conclusion}
%% -----------------------------------------------------------------------

SysPlug provides a practical, zero-cost path from model and hardware
specification to an optimised, memory-safe training configuration.  Its
combination of analytic memory prediction, roofline throughput estimation,
and online stability monitoring addresses the most common failure modes in
practical LLM training.  We release SysPlug as an open-source library under
the MIT license.

\bibliographystyle{plain}
\bibliography{refs}

\appendix

%% -----------------------------------------------------------------------
\section{Memory Model Derivation}
\label{app:memory}
%% -----------------------------------------------------------------------

\subsection{Activation Memory Formula}

A standard transformer block with hidden dimension $H$ computes:
\begin{itemize}
  \item QKV projections: $3 \times B \times S \times H$ activations
  \item Attention scores: $B \times S \times S$ (full attention) or $O(S)$ with FlashAttention
  \item Attention output: $B \times S \times H$
  \item FFN: $2 \times B \times S \times 4H$ (two projections)
  \item Residual connections: $2 \times B \times S \times H$
  \item LayerNorm: $2 \times B \times S \times H$
\end{itemize}

Summing and simplifying gives approximately $34 \times B \times S \times H$
elements per layer, matching our formula.

\subsection{Gradient Checkpointing Reduction}

With gradient checkpointing, only $\sqrt{L}$ evenly-spaced activation
checkpoints are stored across $L$ layers.  Each forward pass between
checkpoints is recomputed during the backward pass.  The storage factor
becomes $\sqrt{L}/L = 1/\sqrt{L}$, reducing from $O(L)$ to $O(\sqrt{L})$.

%% -----------------------------------------------------------------------
\section{Full API Reference}
\label{app:api}
%% -----------------------------------------------------------------------

\begin{lstlisting}[language=Python, caption={Core SysPlug API}]
import sysplug

# Create advisor
advisor = sysplug.Advisor(
    model,                  # nn.Module | str | int
    hardware="auto",        # "auto" | HardwareSnapshot
    training_type="sft",    # supervised|sft|dpo|rlhf|grpo
    objective="balanced",   # throughput|memory|balanced
    verbose=True,
)

# Suggest configuration
cfg = advisor.suggest_config({
    "batch_size": 8,
    "learning_rate": 2e-5,
    "precision": "bf16",
})
# cfg: SysPlugConfig

# What-if analysis
result = advisor.what_if({"batch_size": 32})
# result.new_config, .changed_params, .reason, .feasible

# Online monitoring
with advisor.monitor(check_interval_steps=100) as mon:
    for step, batch in loader:
        loss = train_step(batch)
        mon.record(step=step, loss=loss.item())

# Framework integrations
ta = cfg.to_training_arguments(output_dir="./out")
ds = cfg.to_deepspeed_config()
cfg.apply_to_optimizer(optimizer)
\end{lstlisting}

\end{document}
